---
title: "Getting Started with PriorPostPlots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PriorPostPlots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(PriorPostPlots)
```

## Why check priors against posteriors?

Every Bayesian model has priors, and it's worth asking, for each parameter: did the data actually
tell you something, or is the answer mostly just your prior talking back to you? `PriorPostPlots`
answers that visually. Give it a fitted Stan model and the Stan code it came from, and it draws one
panel per parameter: a shaded curve for the prior, and a histogram of the posterior draws, on the
same axis.

The whole package is really one function: **`PriorPosteriorPlot()`**. It reads the priors and hard
bounds straight out of your Stan code, and the posterior draws straight out of your fitted model -
there's no separate data frame to build by hand.

This tutorial builds up from a single parameter to a small full example, using *simulated* posterior
draws throughout (via a tiny stand-in for a fitted model), so every chunk here runs instantly with
no Stan installation required. The last section shows what the equivalent code looks like against a
model you've actually fitted.

## A single parameter

Suppose a model has one parameter, `mu`, with a fairly wide `normal(0, 10)` prior:

```{r}
stan_code <- "
parameters {
  real mu;
}
model {
  mu ~ normal(0, 10);
}
"
```

Now suppose the data turned out to be quite informative, and the posterior for `mu` ended up
concentrated around 5, with a small spread. We'll simulate that directly rather than actually
fitting anything:

```{r}
set.seed(1)
draws <- matrix(rnorm(1000, mean = 5, sd = 0.4), ncol = 1, dimnames = list(NULL, "mu"))
```

`PriorPosteriorPlot()` needs something that *behaves* like a fitted Stan model - specifically,
something with a `$draws(variables, format = "matrix")` method (this is exactly what a real
`cmdstanr` fit provides). A plain list works fine as a stand-in:

```{r}
mock_fit <- list(draws = function(variables, format = "matrix") draws[, variables, drop = FALSE])

PriorPosteriorPlot(mock_fit, stan_code)
```

## Reading the plot

- **Histogram narrow and off to one side of the shaded region**: the data pulled the estimate away
  from the prior's own center, and did so with some precision. That's the case above - the prior was
  centered at 0 and fairly spread out, but the posterior landed confidently around 5.
- **Histogram spread out and similar in shape/width to the shaded prior region**: harder to tell
  whether the data added much, since the posterior looks like the prior. This isn't necessarily a
  problem - it can just mean the prior was already a reasonable summary, or that this particular
  parameter is only weakly identified by the data you have. It's a prompt to look closer, not an
  automatic verdict.
- **Histogram pushed right up against the edge of the shaded region**: worth a second look at
  whether the prior's range was too narrow for what the data actually wants to say. `PriorPosteriorPlot()`
  always widens the display so every posterior draw is visible on the same axis as the prior, so this
  shows up as the ribbon extending unusually far to reach the histogram, rather than the histogram
  being cut off.

## Multiple parameters, including a vector

Real models have more than one parameter, and vector-valued parameters are common. `pars` accepts
a *base* name for a vector and expands it to every element automatically - you never need to list
`beta[1]`, `beta[2]`, ... by hand.

```{r}
stan_code <- "
parameters {
  real mu;
  vector[2] beta;
}
model {
  mu ~ normal(0, 5);
  beta ~ student_t(4, 0, 2);
}
"

set.seed(2)
draws <- cbind(
  mu        = rnorm(1000, 1.8, 0.3),
  `beta[1]` = rnorm(1000, 0, 1.9),    # close to its prior - only weakly informed
  `beta[2]` = rnorm(1000, -3, 0.4)    # far from its prior - strongly informed
)
mock_fit <- list(draws = function(variables, format = "matrix") {
  # a real cmdstanr fit expands a base name like "beta" to "beta[1]", "beta[2]", ...
  # automatically; this mock has to do that matching itself
  cols <- unlist(lapply(variables, function(v) grep(paste0("^", v, "(\\[|$)"), colnames(draws), value = TRUE)))
  draws[, cols, drop = FALSE]
})

PriorPosteriorPlot(mock_fit, stan_code)
```

Note the contrast between the two `beta` panels: `beta[1]`'s posterior looks a lot like its prior
(weakly informed), while `beta[2]`'s has moved decisively away from 0 (strongly informed) - exactly
the kind of per-element difference that's easy to miss without plotting each one.

## A bounded parameter

Variance/scale parameters are usually given a half-normal or half-Cauchy prior - a distribution with
support restricted to positive values via a `<lower=0>` bound. This is detected automatically and
drawn as a genuine cliff-edge, not just a display cutoff:

```{r}
stan_code <- "
parameters {
  real<lower=0> sigma;
}
model {
  sigma ~ cauchy(0, 1);
}
"

set.seed(3)
draws <- matrix(abs(rnorm(1000, 0.3, 0.15)), ncol = 1, dimnames = list(NULL, "sigma"))
mock_fit <- list(draws = function(variables, format = "matrix") draws[, variables, drop = FALSE])

PriorPosteriorPlot(mock_fit, stan_code)
```

## When you need `data_list`

Stan models often size a `vector` or `matrix` using a named constant declared in the `data` block,
rather than a literal number:

```stan
data {
  int<lower=1> K;
}
parameters {
  vector[K] beta;
}
```

`PriorPosteriorPlot()` only ever reads the Stan *code* as text, so on its own it has no way to know
what `K` actually equals. There are two ways it can find out. The first is `data_list`: pass it the
same named list of data you gave to `mod$sample(data = ...)` (or `rstan::stan(data = ...)`), and any
bound or prior argument written as a name rather than a literal number is resolved from it. It
doesn't need to be the *complete* data list you fit with - just enough of it to resolve whatever
named constants the code actually references.

```{r}
stan_code <- "
data {
  int<lower=1> K;
}
parameters {
  vector[K] beta;
}
model {
  beta ~ normal(0, 1);
}
"

set.seed(4)
draws <- matrix(rnorm(3000, 0, 1), ncol = 3, dimnames = list(NULL, paste0("beta[", 1:3, "]")))
mock_fit <- list(draws = function(variables, format = "matrix") {
  cols <- unlist(lapply(variables, function(v) grep(paste0("^", v, "(\\[|$)"), colnames(draws), value = TRUE)))
  draws[, cols, drop = FALSE]
})

PriorPosteriorPlot(mock_fit, stan_code, data_list = list(K = 3))
```

The second way is automatic, and needs nothing from you: if `data_list` is left out (or doesn't
include the constant needed), `PriorPosteriorPlot()` falls back to inferring the size from
`stan_fit`'s own posterior draws - Stan itself had to resolve `K` to produce them, so a `beta[1]`,
`beta[2]`, `beta[3]` in the fit is just as good a source for "`beta` has 3 elements" as `data_list`
would have been. In fact, the call above would produce the same three panels even with
`data_list` left out entirely, since `mock_fit` already exposes all three `beta[...]` columns. This
fallback only fires once resolving the size from a literal or `data_list` has already failed, so
passing `data_list` never hurts and always takes priority when both are available.

What this fallback *can't* do is resolve a value that isn't part of the fitted output at all - a
named constant used as a **prior's own argument**, rather than as a size:

```{r}
stan_code <- "
data {
  real mu_prior;
}
parameters {
  real beta;
}
model {
  beta ~ normal(mu_prior, 1);
}
"

set.seed(5)
draws <- matrix(rnorm(1000, 2, 0.5), ncol = 1, dimnames = list(NULL, "beta"))
mock_fit <- list(draws = function(variables, format = "matrix") draws[, variables, drop = FALSE])

PriorPosteriorPlot(mock_fit, stan_code)  # no data_list - mu_prior can't be resolved
```

`mu_prior` never appears in the posterior draws - it's an input to the prior, not an output of
fitting - so there's no fallback for it. `beta`'s panel is still drawn (the parameter itself isn't
dropped), but with no prior ribbon, since the prior's center couldn't be pinned down. Passing
`data_list = list(mu_prior = 0)` resolves it and restores the ribbon. This is the one situation
`data_list` remains genuinely necessary for: a prior argument that names a `data`-block constant
directly, rather than a `vector`/`matrix` size.

Either way - an unresolved size or an unresolved prior argument - a warning names exactly which
parameter and which constant it couldn't resolve, so it's easy to spot.

## Supported priors, at a glance

| Stan syntax | Notes |
|---|---|
| `normal(mu, sigma)` | |
| `lognormal(mu, sigma)` | |
| `exponential(rate)` | |
| `gamma(shape, rate)` | |
| `uniform(lower, upper)` | |
| `double_exponential(mu, scale)` | the Laplace distribution |
| `beta(shape1, shape2)` | |
| `cauchy(mu, sigma)` | `<lower=0>` renders as half-Cauchy |
| `student_t(nu, mu, sigma)` | |
| `lkj_corr_cholesky(eta)` | on a `cholesky_factor_corr[K]` parameter - see "Common gotchas" below |

Full argument-by-argument detail is in `?PriorPosteriorPlot`.

## Using this with a model you've actually fitted

Everything above used simulated draws so this document runs on its own. Real usage looks like this
(not run here) - a few common ways you might already be generating samples, all of which work
directly with `PriorPosteriorPlot()`:

```{r, eval = FALSE}
# cmdstanr - compile once, then sample. Real fits normally use several
# chains and a warmup phase; PriorPosteriorPlot() doesn't need to know
# about either - it automatically pools every chain's post-warmup draws.
library(cmdstanr)
mod <- cmdstan_model("my_model.stan")
fit <- mod$sample(data = my_data, chains = 4, iter_warmup = 1000, iter_sampling = 1000)
PriorPosteriorPlot(fit, "my_model.stan")

# rstan - compiles and samples in one step (note: rstan's `iter` argument
# is warmup + sampling combined, rather than two separate arguments)
library(rstan)
fit <- stan(file = "my_model.stan", data = my_data, chains = 4, warmup = 1000, iter = 2000)
PriorPosteriorPlot(fit, "my_model.stan")

# already have a fitted object, but not the original file or code string?
# a stanfit already carries its own compiled model, so nothing needs
# extracting - just hand that straight to PriorPosteriorPlot():
PriorPosteriorPlot(fit, fit@stanmodel)

# a cmdstanr fit can hand back its own Stan code the same way:
PriorPosteriorPlot(fit, fit$code())

# restrict to specific parameters (see "When you need data_list" above
# for what a data_list is for and what it needs to contain)
PriorPosteriorPlot(fit, "my_model.stan", pars = c("beta", "sigma"),
                    data_list = my_data)
```

## Common gotchas

- **Correlated random effects.** If your model declares `cholesky_factor_corr[K] L;` with an
  `L ~ lkj_corr_cholesky(eta);` prior (the standard way to give two or more random effects a
  correlation), `PriorPosteriorPlot()` cannot plot `L`'s own raw entries - they aren't individually
  meaningful. It plots the actual pairwise correlation(s) instead, but only if your Stan code
  *also* computes them, via the standard line
  `corr_matrix[K] Corr_L = multiply_lower_tri_self_transpose(L);` in `transformed parameters` or
  `generated quantities`. If that line is missing, you'll get a warning telling you exactly what to
  add - it's not an error, just nothing to plot for that parameter yet.
- **An unrecognised prior gets no ribbon, not an error.** If a parameter's sampling statement uses a
  distribution this package doesn't know (or has none at all, and no hard bounds), its panel simply
  shows the histogram with no shaded region - a visual cue to double check that parameter by hand.
- **Big models produce a lot of panels.** Leaving `pars = NULL` plots every parameter this function
  recognises, which can be a lot for a hierarchical model with many random effects. Passing
  `pars = c(...)` with just the parameters you actually want to check also keeps the console quiet
  about anything unusual in the parts of the model you weren't asking about.
